---
title: "assignment"
output: html_document
date: "2026-04-28"
---

```{r}
suppressPackageStartupMessages({
  library(epiwraps)
  library(ggplot2)
  library(AnnotationHub)
  library(ensembldb)
  library(edgeR)
  library(rtracklayer)
  library(BiocParallel)
  library(bsseq)   
  library(DMRcate)
  library(GenomicRanges)
  library(GenomeInfoDb)
})
```
#SetUp of genomic data
```{r, load bigwigs}
#Load DMR ranges

DNAme <- readRDS("comparison/DNAme.rds")
```

```{r, annotation}
#Get Annotaion
ah <- AnnotationHub()
ensdb <- ah[["AH89211"]] # GRCm38
```
#Task 1
```{r}
#This code bit was not necessary in the end
#Extract the genes
##Only use chr 1
chr1 <-  GRanges(seqnames=Rle(c("1")), 
                 ranges = IRanges(1, end=195471971))

genes_chr1 <- genes(ensdb, filter=GRangesFilter(chr1), columns=c("gene_name"))

head(genes_chr1)
```

#Differential Methylation Testing
```{r}
#Extract the methylation data from the BSseq object from ESC's and astrocytes

bs <- readRDS("ESC_astrocytes_BSseq.rds")
## remove CpGs with 0 coverage
bs <- bs[rowSums(assay(bs,"Cov")>0)==ncol(bs),]
bs <- bs[seqnames(bs) == "chr1",]
design <- model.matrix(~celltype, data=pData(bs)) 
methdesign <- modelMatrixMeth(design)
```

```{r}
genes <- readRDS("hg19_chr1_genes.rds")
head(genes$symbol)  
```

```{r}
#find the top 5 regions
seqAnnot <- sequencing.annotate(bs, methdesign, all.cov=TRUE, coef="celltypeastrocytes")
dmrcateRes <- dmrcate(seqAnnot, C=2, min.cpgs=5)
dmrRanges <- extractRanges(dmrcateRes, genome="hg19")

top5 <- dmrRanges[order(abs(dmrRanges$meandiff), decreasing=TRUE)][1:5]
seqlevels(top5) <- "chr1"

# For verfification and analysis
as.character(seqnames(top5))  
head(top5)
```

```{r}
#Save relevant gene names in a vector
gene_names <- c("CR1L", "CD46P1", "C1orf198", "RP11-115D7.3")
dmr_genes <- genes[genes$symbol %in% gene_names]

# Extract methylation
metDMR <- bsseq::getMeth(bs, regions=dmr_genes, what="perRegion", type="raw")
rownames(metDMR) <- dmr_genes$symbol
colnames(metDMR) <- colnames(bs)

print(metDMR)
pheatmap(metDMR, name="DMR\nmethylation", show_rownames=TRUE, fontsize_row=8)

```

#Task 2
##Unfortunately I was not able to run the code snce it would need SeInfo which is contained in the library(GenomeInfoDb), which is a library not compatible with my R version. I was not able to find any other solution or workarounds. SeqInfo what have been needed to read this .rds file with the ATAC data. AS a result I wrote the whole code in theory, however I was not able to test it.
```{r}
#download.file("https://www.ethz-ins.org/content/ATAC_AstroVsESC_subset.rds", destfile = "accessibility_data.rds")
#atac <- readRDS("accessibility_data.rds")
#dmrRanges_new<- readRDS("dmrRanges_new.rds")
```

```{r}
# Find overlaps between DMRs and ATAC peaks
#hits2 <- findOverlaps(dmrRanges_new, atac)

# Extract matched pairs
#dmr_matched <- dmrRanges_new[queryHits(hits2)]
#atac_matched <- atac[subjectHits(hits2)]

# dataframe with both signals
#df <- data.frame(
 # methylation_diff = dmr_matched$meandiff,      # from DMRcate
 # atac_log2FC = atac_matched$log2FoldChange     # from ATAC - column name may differ!
#)

#Theoretical interpretation of the plot: if methylation decreases at in a region=accessibility should increase (meaning mean methylation difference is bigger) in these reagions than in others
#ggplot(df, aes(x=methylation_diff, y=atac_log2FC)) +
 # geom_point() +
 # geom_smooth(method="lm") +
 # geom_hline(yintercept=0, linetype="dashed") +
 # geom_vline(xintercept=0, linetype="dashed") +
 # labs(x="Mean methylation difference (astrocyte - ESC)",
  #     y="ATAC log2 fold change (astrocyte vs ESC)",
  #     title="Agreement between methylation and accessibility changes")
```