---
title: "Week04 part 2 : Comparing multiple signals"
author: "Pierre-Luc"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(ensembldb)
  library(GenomicRanges)
  library(epiwraps)
  library(rtracklayer)
  library(ggplot2)
})
# we shouldn't actually need this today, but in case:
# ah <- AnnotationHub(localHub=TRUE)
# ensdb <- ah[["AH89211"]]
```

# Comparing multiple signals

## Downloading the data

Download peaks and signals for in mESC, reduced to chr1 to make things more lightweight:

```{r, eval=FALSE}
options(timeout=1800)
download.file("https://ethz-ins.org/content/p300_subset.zip", "p300_subset.zip")
unzip("p300_subset.zip")
```

## Overlaps between peaks

```{r}
# to read a single peak file:
# peaks <- rtracklayer::import("peaks/H3K27ac.bed", format="narrowPeak")
# to read several peak files:
# first list the files:
peakfiles <- list.files("peaks", full.names = TRUE)
# then give them meaningful names
names(peakfiles) <- gsub("\\.bed","",basename(peakfiles))
# then loop through them and read them:
peaks <- lapply(peakfiles, FUN=rtracklayer::import, format="narrowPeak")
```

Check, for each set of peaks, what proportion overlaps with the p300 peaks:

```{r}
sapply(peaks, FUN=function(x) sum(overlapsAny(peaks$p300, x))/length(peaks$p300))
```

Here are other ways of visualizing overlaps between peaks

```{r}
# As a heatmap of overlaps:
regionOverlaps(peaks)
# As an upset plot:
tmp <- regionsToUpset(peaks, nsets=length(peaks))
UpSet(make_comb_mat(regionsToUpset(peaks, nsets=length(peaks))))
```

By default, this is comparing each set of peaks to a reference that is made by the (merged) union of all peaks. However, we can also compared to a pre-defined reference:

```{r}
ComplexHeatmap::UpSet(make_comb_mat(regionsToUpset(peaks, reference = peaks$p300, nsets=length(peaks))))
```

## Visualizing the p300 regions

```{r}
# list the bigwig files
bwfiles <- list.files("tracks", full.names = TRUE)
# then give them meaningful names
names(bwfiles) <- gsub("\\.bw","",basename(bwfiles))

# read the signal around the p300 peaks:
sm <- signal2Matrix(bwfiles, regions=peaks$p300, w=50)
# we could plot them all on separate scales:
plotEnrichedHeatmaps(sm)
# when doing that it can also help to use different colors, e.g.:
plotEnrichedHeatmaps(sm, trim = 0.95, multiScale = TRUE, axis_name = "0", 
                     colors=list("darkred","yellow","black","darkgreen","darkblue",
                                 viridisLite::inferno(100)))
```

We can also try a quick approximative normalization method, to put everything on a comparable scale:

```{r}
sm <- renormalizeSignalMatrices(sm, method="top", trim = 0.05)
plotEnrichedHeatmaps(sm, trim=0.95)
```


# Looking up at exclusive peaks

## K27ac but not p300

```{r}
k27acOnly <- peaks$H3K27ac[!overlapsAny(peaks$H3K27ac,peaks$p300)]

sm2 <- signal2Matrix(bwfiles[c(1,5)], regions=k27acOnly, w=50)

plotEnrichedHeatmaps(sm2)
```

## p300 but not K27ac

```{r}
p300Only <- peaks$p300[!overlapsAny(peaks$p300,peaks$H3K27ac)]

sm.p300 <- signal2Matrix(bwfiles[c(1,5)], regions=p300Only, w=50)

plotEnrichedHeatmaps(sm.p300)
```



Now let's have a look at the union of all peaks:

```{r}
unionPeaks <- reduce(unlist(GRangesList(peaks)))
sm <- signal2Matrix(bwfiles, regions=unionPeaks, w=50)
sm2 <- renormalizeSignalMatrices(sm, method="top")
plotEnrichedHeatmaps(sm2, trim=0.95) # same scale
plotEnrichedHeatmaps(sm, trim=0.95, multiScale = TRUE) # separate scales
```

