---
title: "Motif accessibility analysis using betterChromVAR"
author: "Pierre-Luc"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(motifmatchr)
  library(MotifDb)
  library(universalmotif)
  library(ggplot2)
  library(sechm)
  library(BSgenome.Mmusculus.UCSC.mm10)
  library(limma)
  library(betterChromVAR)
})

ah <- AnnotationHub(localHub=TRUE)
ensdb <- ah[["AH89211"]] # mouse ensembldb object
```

## Getting 

## Download the data

For this practical, we are going to re-use the hippocampus ATAC-seq data from week 5:

```{r, eval=FALSE}
download.file("https://ethz-ins.org/content/mouse_mm38_hippocampus.peakCounts.SE.rds", "peakCounts.rds", mode="wb")
```

```{r}
se <- readRDS("peakCounts.rds")
```

The object contains fragment counts in peaks across control and stressed samples.

To build such an object, one can use the `peakCountsFromBAM` of `epiwraps`.
Note that for what follows, it's important that the regions be similarly-sized.
Before getting the counts, it's therefore recommended to resize the regions, e.g. 
doing `resize(peaks, width=300, fix="center")`.

## Annotate regions

The first thing we need to do is figure out the GC content of the regions, as well as what motif matches each region. The first is easily done:

```{r}
# we restrict the regions to standard chromosomes and change the seq level style to match the genome
se <- GenomeInfoDb::keepStandardChromosomes(se, pruning.mode = "coarse")
seqlevelsStyle(se) <- "UCSC"
se <- addGCBias(se, genome=BSgenome.Mmusculus.UCSC.mm10)
# (to use a fasta file as genome, we could simply do genome=FaFile("path/to/genome.fasta") )
```

### Motif matches

For motif annotations, we first need to get a list of motifs, and then scan the regions for matches to them.
We could fetch motifs from the `MotifDb` package:

```{r}
# make sure you specify a collection, in order to avoid having redundant motifs:
motifs <- query(MotifDb, "HOCOMOCOv13")
```

Alternatively, we could get the motifs from whatever source we like. What we often like to use in such analyses are 'archetypes', or non-redundant motifs representing clusters of TFs. We can for instance get those from HOCOMOCO v.14:

```{r, eval=FALSE}
download.file("https://hocomoco14.autosome.org/final_bundle/hocomoco14/H14CORE-CLUSTERED/H14CORE-CLUSTERED_pwm.tar.gz", "H14CORE_clusters.tar.gz")
untar("H14CORE_clusters.tar.gz")
lf <- list.files("H14CORE-CLUSTERED/pwm/", full=TRUE)
archetypes_motifs <- lapply(lf, read_matrix, headers = ">", positions = "rows")
```

Either way, we have to convert the motifs to something that `motifmatchr` accepts:

```{r, message=FALSE}
motifs <- do.call(TFBSTools::PWMatrixList, convert_motifs(motifs, class="TFBSTools-PWMatrix"))
names(motifs) <- make.unique(sapply(motifs, \(x) gsub("\\..+","",x@name)))
```

We can then scan regions for matches of the motifs:

```{r}
matches <- motifmatchr::matchMotifs(motifs, se, genome=BSgenome.Mmusculus.UCSC.mm10)
matches
```

The results is also a SummarizedExperiment, where columns are the motifs and the rows are the regions, and the assay is a logical sparse matrix:

```{r}
assay(matches)[1:10,1:20]
```

## Get motifs deviations

```{r}
dev <- betterChromVAR(se, matches)
dev
```

The output is also a SummarizedExperiment, this time with samples as columns and motifs as rows.
The values represent the relative accessibility of each motif in each sample.

## Differential analysis

We can use these motif deviations to perform differential analysis:

```{r}
dev$group <- ifelse(grepl("FSS", colnames(dev)), "Stressed", "Control")
design <- model.matrix(~group, data=as.data.frame(colData(dev)))
fit <- eBayes(lmFit(assay(dev, "z"), design))
res <- as.data.frame(topTable(fit, number = Inf))
res$motif <- row.names(res)
ggplot(res, aes(logFC, -log10(adj.P.Val), label=motif)) + geom_text()
```

We could visualize the top hits:

```{r}
sechm(dev, head(row.names(res)), gaps_at="group", assayName = "z")
```

